Accurate binding of sodium and calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-11-14 Manuscript/manuscript.tex
Small edits from II B to III B

The relation between the amount of the bound charge per lipid, $X^\pm$, and the head group order parameter change, $\Delta S_{\rm{CH}}^{i}$, is empirically quantified as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx m_i \frac{4 }{3\chi}X^\pm,
\end{equation}
where $i$ refers to either $\alpha$ or $\beta$ carbon, $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge, $\chi$ is the quadrupole coupling constant ($\chi \approx$\,167\,kHz), and $m_i$ empirical experimental constant depending on the valency and position of the bound charge.

The measured change of the order parameter depends both on the head group response to the bound charge and on the amount of the bound charge (\textit{i.e.,} $m_i$ and $X^\pm$ in Eq.~\ref{OPchangeEQ}, respectively).  The empirical factor $m_i$ has to be adequately quantified before the electrometer concept can be used to analyze the binding affinities. This calibration has been done experimentally for a wide range of systems~\cite{seelig87, beschiasvili91}. To calibrate the response of the head group order parameter to the bound charge in simulations, we use experimental data for a strong cationic surfactant dihexadecyldimethylammoniumbromide  (DHAB) bilayer~\cite{scherer89}. DHAB\\[0.5cm]
%\vspace{0.5cm}
\chemfig{
  -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH_2{)}_{15}-CH_3)(-[:30]{(}CH_2{)}_{15}-CH_3)
}  
\vspace{0.5cm} \\
is a cationic surfactant having two acyl chains and bearing a unit charge in the hydrophilic end. Thus, it is expected to locate in the bilayer similarly to the phospholipids, and its molar ratio then gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems~\cite{scherer89}.

\subsection{Salt concentrations and binding affinity}
NMR experiments report the used salt concentrations in two different ways when measuring head group order parameters. Some use the salt concentrations in water before solvating the lipids~\cite{akutsu81} while others use atomic absorption spectroscopy and report the salt concentration in the supernatant after the solvation of lipids~\cite{altenbach84}. In this work, we use the latter definition. In particular, we measure the salt concentration in the aqueous bulk region as that at the farthest point from the lipid bilayer in the aqueous phase. Notice that this definition differs from the one used by Catte et al.~\cite{catte16} where the former definition is used. Although these two definitions vary when applied to Ca$^{2+}$, they do not affect significantly the conclusions of this work.

To quantify the ion binding affinities to a lipid bilayer, we calculate the relative surface excess of ions with respect to water, $\Gamma_{ion}^{water}$~\cite{chattorajBOOK}. Such a quantity compares the adsorption of ions to the adsorption of water molecules at the interface without the necessity of defining a Gibbs dividing plane between the membrane interior and the water bulk region. In our simulations, we only assume that the interface is located between the ion-free hydrophobic interior of the lipid bilayer and the aqueous region far from the membrane.  Such a setup and the above definition of bulk ion concentration provide a simple relation for the relative surface excess $\Gamma_{ion}^{water}$ for simulations of lipid bilayers,

\begin{equation}\label{surfexcess}
  \Gamma_{ion}^{water}=\frac{1}{2A_b} \left ( n_{ion} - n_{water} \frac{C_{ion}}{C_{water}} \right ).
\end{equation}
Here, $n_{water}$ and $n_{ion}$ are the total numbers of water molecules and ions in the system,
$C_{water}$ and $C_{ion}$ are their respective bulk concentrations in the aqueous phase,
and $A_b$ is the area of the unit cell in the membrane plane.
The total area of the interface is then twice the area of the membrane, i.e., $2A_b$,
because the bilayer has an interface at each of the two leaflets.

@@

\subsubsection{Simulations of POPC bilayers with aqueous ions}
Simulations of a POPC bilayer in pure water or at varying salt concentrations consisted of 128 POPC molecules and approximately 50~SPC/E~water~\cite{Berendsen1987} molecules per lipid in an orthorhombic simulation box with periodic boundary conditions.  The chosen SPC/E water model, also used in the previous parametrization of ECC-ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}, has a lowered dielectric constant, consistent with the ECC concept~\cite{leontyev11, leontyev14}.  We checked other water models (OPC \cite{Izadi14}, OPC3~\cite{Izadi16}, TIP3P~\cite{jorgensen83}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005}) to confirm the robustness of our approach (See SI). Sodium, calcium, and chloride ions were modeled as ECC-ions with parameters from Ref. \citenum{jungwirth17-new-paper-to-be-published, Pluharova2014}.  
% cite order: [Ca_2s, Na_s, Cl_2s] 
Scaled charges and van der Waals radii for atoms forming the POPC lipid are derived in this work starting from the Lipid14 force field~\cite{REF}. For comparison, simulations with unscaled charges were performed using the Lipid14 model~\cite{REF}, the TIP3p model for water \cite{jorgensen83}\todo{HECTOR: Do we compare this against scaled with SPC or TIP3P?}, and ion models by Dang and coworkers~\cite{smith94,chang1999,dang2006}. Simulation data for the Lipid14 model with \AA{}qvist ions~\cite{aqvist90} and TIP3P~\cite{jorgensen83} water models were taken directly from Ref.~\citenum{catte16}.

@@

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
The topology of dihexadecyldimethylammonium was created with the automated topology builder \cite{malde11}. The AmberTools program~\cite{amber} was then used to generate the Amber-type force field parameters. These parameters were then converted to the Gromacs format with the acpype tool \cite{acpype}. The partial charges were then manually modified in the non-scaled version~\cite{catte16} to approximately match similar segments in Lipid14 \cite{dickson14}. For simulation with ECC-POPC, the partial charges of the headgroup were scaled, i.e., $f_q=0.8$, and the $\sigma$ of the scale atoms reduced as in POPC-ECC.

The cationic surfactants were randomly mixed among the phospholipids to form bilayer structures with mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer. All these systems contained 50 POPC molecules per leaflet, 6340 TIP3P~\cite{jorgensen83} water molecules, and 6, 14, 21, 35, or 50 surfactants per leaflet.  Chloride counter ions were used in simulations, while bromide was the counter ion in the experiment \cite{scherer89}. For POPC either the Lipid14 or ECC-POPC model was used. The first 20~ns of the total simulation time of 200~ns was considered as an equilibration time and was omitted from the analysis. Sufficient lipid neighbor exchange occurred during the simulations. Further details for the unscaled simulations and the trajectories are in Ref.~\citenum{catte16}.

@@

\begin{table}[tb!]
  \caption{Values of the area per lipid (APL) of POPC bilayers without ions from the Lipid14 simulation
    ran in this work and from the literature \cite{dickson14}, from the ECC-lipids model, and from experiments.{ \color{red} HECTOR:Are we using here the same water model? If not is it fair the comparison considering the large changes observed in the SI table?}\label{tab:apls} }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14                   & 65.1$\pm$ 0.6  &  300 \\
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids                & {\color{red}63.2}$\pm$ 0.6  &  300       \\
    \hline
    experiment \cite{kucerka11} & 64.3  &  303    \\
    \hline
  \end{tabular}
\end{table}

First, we present results for bilayers in pure water.
The ECC-lipids and Lipid14 models both reproduce the experimental x-ray scattering form factors
of a POPC bilayer with a comparable accuracy in Fig.~\ref{simVSexpNOions}.
The area per lipid from the Lipid14 model is $\approx$1\AA{} larger than the
experimental value in Table~\ref{tab:apls}, while the value from the ECC-lipids model
is $\approx$1\AA{} smaller\todo{The 63.2 value does not appear in the SI table. There it is 62.2 but I'm not sure corresponds to the same system}. The values of the area per lipid of the ECC-lipids model show larger variations
when simulated with different water models (i.e., 62.2--66.8 \AA{}, see Table~\ref{tab:apls_si} in SI\todo{HECTOR: change to real reference in SI}),
while still being close to the experimentally reported values.
We can thus conclude that the ECC-lipids model reproduces the experimental dimensions of the POPC
lipid bilayer with accuracy comparable to that for another state of the art lipid models \cite{ollila16}.
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

@@

Before studying the sodium and calcium ion binding affinities, we have to quantify as reference the response of the headgroup order parameters to the amount of bound charge by using mixtures of monovalent cationic surfactants (dihexadecyldimethylammonium) and POPC~\cite{scherer89}. These mixtures have a well-defined amount of bound charge per PC, i.e., the molar fraction of cationic surfactants. This direct relation results from the ability of dihexadecyldimethylammonium to be inserted in the lipid bilayers as any lipid because of its two hydrophobic acyl chains. Furthermore, available experimental data for these systems can be used to validate the sensitivity of lipid headgroup order parameters to the amount of bound charge in simulations \cite{REF}.

The changes of the headgroup order parameters with increasing amount of the cationic surfactant are compared between simulations and experiments~\cite{scherer89} in Fig.~\ref{OrderParameterCHANGESsurf}. An approximately linear decrease of the order parameters, as expected from Eq. \ref{OPchangeEQ}, is observed in both simulations and experiments at least for mole fractions below $\sim$30\%. The slope is, however, too steep in the Lipid14 model indicating that the head group order parameters response is too large to the bound positive charge. In contrast, the slope of the ECC-lipids model is in a very good agreement with experiments for the $\alpha$ segment, while being slightly underestimated for the $\beta$ segment.

In Fig.~\ref{OrderParameterCHANGESsurf}, we show the headgroup P-N vector angle as a function of the mole fraction of the cationic surfactant. As suggested previously~\cite{seelig87}, the headgroup orients more towards the water phase with the increasing amount of positive charge in a PC lipid bilayer. The effect is more pronounced in the Lipid14 model than in the ECC-POPC model.   For example, the addition of 50\% mole fraction of the cationic surfactant leads to the decrease of 20$^{\circ}$ of the P-N vector angle for the Lipid14 model while only 11$^{\circ}$ in the ECC-POPC model. The difference is in line with the smaller order parameter changes and the reduced charge--dipole interactions in the latter model. The lower sensitivity of the P-N vector angle response in the ECC-POPC model agrees better with experiments. The results imply that improvements are required in MD models for reliable studies of the lipid headgroup responses to ions or other biomolecules{\color{red}, as also suggested previously \cite{botan15}} \todo{Do you want to finish each paragraph with a citation to your previous works?}.

>>>>

The relation between the amount of the bound charge per lipid, $X^\pm$, and the head group order parameter change, $\Delta S_{\rm{CH}}^{i}$, is empirically quantified as~\cite{seelig87,ferreira16}
\begin{equation}\label{OPchangeEQ}
\Delta S_{\rm{CH}}^{i}= S_{\rm{CH}}^{i}(X^\pm)-S_{\rm{CH}}^{i}(0) \approx m_i \frac{4 }{3\chi}X^\pm,
\end{equation}
where $i$ refers to either $\alpha$ or $\beta$ carbon, $S_{\rm{CH}}^{i}(0)$ denote the order parameter in the absence of bound charge, $\chi$ is the quadrupole coupling constant ($\chi \approx$\,167\,kHz), and $m_i$ is a empirical constant depending on the valency and position of the bound charge.

The measured change of the order parameter depends both on the head group response to the bound charge and on the amount of the bound charge (\textit{i.e.,} $m_i$ and $X^\pm$ in Eq.~\ref{OPchangeEQ}, respectively).  The empirical factor $m_i$ has to be adequately quantified before the electrometer concept can be used to analyze the binding affinities. This calibration has been done experimentally for a wide range of systems~\cite{seelig87, beschiasvili91}. To calibrate the response of the head group order parameter to the bound charge in simulations, we use experimental data for a strong cationic surfactant dihexadecyldimethylammoniumbromide  (DHAB) mixed with a POPC bilayer~\cite{scherer89}. DHAB\\[0.5cm]
%\vspace{0.5cm}
\chemfig{
  -[:0,3.5,,,draw=none]\chemabove{N}{\scriptstyle\oplus} (-[:150]H_3C)(-[:210]H_3C)(-[:330]{(}CH_2{)}_{15}-CH_3)(-[:30]{(}CH_2{)}_{15}-CH_3)
}  
\vspace{0.5cm} \\
is a cationic surfactant having two acyl chains and bearing a unit charge in the hydrophilic end. Thus, it is expected to locate in the bilayer similarly to the phospholipids, and its molar ratio then gives directly the amount of bound unit charge per lipid $X^\pm$ in these systems~\cite{scherer89}.

\subsection{Salt concentrations and binding affinity}
NMR experiments report the used salt concentrations in two different ways when measuring head group order parameters. Some use the salt concentrations in water before solvating the lipids~\cite{akutsu81} while others use atomic absorption spectroscopy and report the salt concentration in the supernatant after the solvation of lipids~\cite{altenbach84}. In this work, we use the latter definition. In particular, we measure the salt concentration in the aqueous bulk region using the farthest point from the lipid bilayer in the aqueous phase. Notice that this definition differs from the one used by Catte et al.~\cite{catte16} where the former definition is used. Although these two definitions vary when applied to Ca$^{2+}$, they do not significantly affect the conclusions of this work.

To quantify the ion binding affinities to a lipid bilayer, we calculate the relative surface excess of ions with respect to water, $\Gamma_{ion}^{water}$~\cite{chattorajBOOK}. Such a quantity compares the adsorption of ions to the adsorption of water molecules at the interface without the necessity of defining a Gibbs dividing plane between the membrane interior and the water bulk region. In our simulations, we only assume that the interface is located between the ion-free hydrophobic interior of the lipid bilayer and the aqueous region far from the membrane.  Such a setup and the above definition of bulk ion concentration provide a simple relation for the relative surface excess $\Gamma_{ion}^{water}$ for simulations of lipid bilayers,
 \begin{equation}\label{surfexcess}
  \Gamma_{ion}^{water}=\frac{1}{2A_b} \left ( n_{ion} - n_{water} \frac{C_{ion}}{C_{water}} \right ).
\end{equation}
Here, $n_{water}$ and $n_{ion}$ are the total numbers of water molecules and ions in the system,
$C_{water}$ and $C_{ion}$ are their respective bulk concentrations in the aqueous phase,
and $A_b$ is the area of the unit cell in the membrane plane.
The total area of the interface is then twice the area of the membrane, i.e., $2A_b$,
because the bilayer has an interface at each of the two leaflets.

@@

\subsubsection{Simulations of POPC bilayers with aqueous ions}
Simulations of a POPC bilayer in pure water or at varying salt concentrations consisted of 128 POPC molecules and approximately 50~SPC/E~water~\cite{Berendsen1987} molecules per lipid in an orthorhombic simulation box with periodic boundary conditions.  The chosen SPC/E water model, also used in the previous parametrization of ECC-ions \cite{jungwirth17-new-paper-to-be-published, kohagen16, Pluharova2014}, has a lowered dielectric constant, consistent with the ECC concept~\cite{leontyev11, leontyev14}.
We tested also other water models (OPC \cite{Izadi14}, OPC3~\cite{Izadi16}, TIP3P~\cite{jorgensen83}, TIP3p-FB and TIP4p-FB~\cite{Wang2014}, and TIP4p/2005 \cite{Abascal2005}) to evaluate the robustness of our approach (See SI). Sodium, calcium, and chloride ions were modeled as ECC-ions with parameters from Ref. \citenum{jungwirth17-new-paper-to-be-published, Pluharova2014}.  
% cite order: [Ca_2s, Na_s, Cl_2s] 
Scaled charges and van der Waals radii for atoms forming the POPC lipid are derived in this work starting from the Lipid14 force field~\cite{dickson14}. For comparison, simulations with unscaled charges were performed using the Lipid14 model~\cite{dickson14}, the TIP3p model for water \cite{jorgensen83}, and ion models by Dang and coworkers~\cite{smith94,chang1999,dang2006}.
\todo{HECTOR: Do we compare this against scaled with SPC or TIP3P? SAMULI: Joe, please fill the information about Lipid14 simulation with ECC-ions.}
Simulation data for the Lipid14 model with \AA{}qvist ions~\cite{aqvist90} and TIP3P~\cite{jorgensen83} water models were taken directly from
Refs.~\citenum{catte16,lipid14POPC0mMNaClfiles,lipid14POPC350mMCaClfiles,lipid14POPC350mMCaClfilesNC,lipid14POPC1000mMNaClfiles}.

@@

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
The topology of dihexadecyldimethylammonium was created with the automated topology builder \cite{malde11}. The AmberTools program~\cite{amber} was then used to generate the Amber-type force field parameters. These parameters were then converted to the Gromacs format with the acpype tool \cite{acpype}. The partial charges were then manually modified in the non-scaled version to approximately match similar segments in Lipid14 \cite{dickson14}. For simulation with ECC-POPC, the partial charges of the headgroup were scaled, i.e., $f_q=0.8$, and the $\sigma$ of the scale atoms reduced as in POPC-ECC.

The cationic surfactants were randomly mixed among the phospholipids to form bilayer structures with mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer. All these systems contained 50 POPC molecules per leaflet, 6340 water molecules, and 6, 14, 21, 35, or 50 surfactants per leaflet.  Chloride counter ions were used in simulations, while bromide was the counter ion in the experiment~\cite{scherer89}. For POPC either the Lipid14 with TIP3P~\cite{jorgensen83} or ECC-POPC with SPC/E~\cite{Berendsen1987}  were used. The first 20~ns of the total simulation time of 200~ns was considered as an equilibration time and was omitted from the analysis. Sufficient lipid neighbor exchange occurred during the simulations. Further details for the unscaled simulations and the trajectories are in Ref.~\citenum{catte16}.

@@

\begin{table}[tb!]
  \caption{Values of the area per lipid (APL) of POPC bilayers without ions from the Lipid14 simulation
    ran in this work and from the literature \cite{dickson14}, from the ECC-lipids model, and from experiments.
    { \color{red} HECTOR:Are we using here the same water model? If not is it fair the comparison considering the large changes observed in the SI table?
      SAMULI: For Lipid14 simulations the water is TIP3P and SPC/E for ECC-lipids. I do understand why the comparison would
    not be fair in this context?}\label{tab:apls} }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14                   & 65.1$\pm$ 0.6  &  300 \\
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids                & {\color{red}63.2}$\pm$ 0.6  &  300       \\
    \hline
    experiment \cite{kucerka11} & 64.3  &  303    \\
    \hline
  \end{tabular}
\end{table}

First, we present results for bilayers in pure water.
The ECC-lipids and Lipid14 models both reproduce the experimental x-ray scattering form factors
of a POPC bilayer with a comparable accuracy in Fig.~\ref{simVSexpNOions}.
The area per lipid from the Lipid14 model is $\approx$1\AA{} larger than the
experimental value in Table~\ref{tab:apls}, while the value from the ECC-lipids model
is $\approx$1\AA{} smaller
\todo{HECTOR: The 63.2 value does not appear in the SI table. There it is 62.2 but I'm not sure corresponds to the same system
SAMULI: I can find the 63.2 from the SI table with SPE/E and the value 62.2 with OPC3, so everything seems to be ok for me.}.
The values of the area per lipid of the ECC-lipids model show larger variations
when simulated with different water models (i.e., 62.2--66.8 \AA{}, see Table~\ref{tab:apls_si} in SI\todo{HECTOR: change to real reference in SI}),
while still being close to the experimentally reported values.
We can thus conclude that the ECC-lipids model reproduces the experimental dimensions of the POPC
lipid bilayer with accuracy comparable to that for nother state of the art lipid models \cite{ollila16}.
%The optimal value for $f_\sigma$ was found by representing the overall membrane structure well
%by matching scattering form factors to experimental data~\cite{Petrache06, Kucerka08, Pabst10}.

@@

Before studying the sodium and calcium ion binding affinities, we have to quantify the response of the headgroup order parameters to the amount of bound charge by using mixtures of monovalent cationic surfactants (dihexadecyldimethylammonium) and POPC~\cite{scherer89}. These mixtures have a well-defined amount of bound charge per PC, i.e., the molar fraction of cationic surfactants. This direct relation results from the ability of dihexadecyldimethylammonium to be inserted in the lipid bilayers as any lipid because of its two hydrophobic acyl chains. Furthermore, available experimental data for these systems can be used to validate the sensitivity of lipid headgroup order parameters to the amount of bound charge in simulations \cite{scherer89}.

The changes of the headgroup order parameters with increasing amount of the cationic surfactant are compared between simulations and experiments~\cite{scherer89} in Fig.~\ref{OrderParameterCHANGESsurf}. An approximately linear decrease of the order parameters, as expected from Eq. \ref{OPchangeEQ}, is observed in both simulations and experiments at least for mole fractions below $\sim$30\%. The slope is, however, too steep in the Lipid14 model indicating that the response of head group order parameters is too large to the bound positive charge. In contrast, the slope of the ECC-lipids model is in a very good agreement with experiments for the $\alpha$ segment, while being slightly underestimated for the $\beta$ segment.

In Fig.~\ref{OrderParameterCHANGESsurf}, we show the headgroup P-N vector angle as a function of the mole fraction of the cationic surfactant. As suggested previously~\cite{seelig87}, the headgroup orients more towards the water phase with the increasing amount of positive charge in a PC lipid bilayer. The effect is more pronounced in the Lipid14 model than in the ECC-POPC model.   For example, the addition of 50\% mole fraction of the cationic surfactant leads to the decrease of 20$^{\circ}$ of the P-N vector angle for the Lipid14 model while only 11$^{\circ}$ in the ECC-POPC model. The difference is in line with the smaller order parameter changes and the reduced charge--dipole interactions in the latter model. The lower sensitivity of the P-N vector angle response in the ECC-POPC model agrees better with experiments. The results imply that improvements are required in MD models for reliable studies of the lipid headgroup responses to ions or other biomolecules.
%{\color{red}., as also suggested previously \cite{botan15}} \todo{Do you want to finish each paragraph with a citation to your previous works?}.
